---
title: "Color scales, color spaces,<br>color-vision deficiency"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: FALSE
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(colorspace)
library(cowplot)
library(ggrepel)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

# color space plot theme
theme_colorspace <- function(size = 12) {
  theme_minimal_grid(size, line_size = 0.2) +
  panel_border(color = "black", size = 0.4) +
  theme(
    axis.ticks.length = grid::unit(0, "pt"),
    axis.ticks = element_line(linewidth = 0.2, color = "black"),
    strip.text = element_text(size = size)
  )
}
```

# Uses of color in data visualization

--

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

---

# Qualitative scale example

.center.nogap[
```{r popgrowth-vs-popsize-colored, echo = FALSE, message = FALSE, out.width = "65%", fig.width = 7.5, fig.asp = 0.618, dev = "svg"}
US_census <- read_csv(here("datasets", "US_census.csv"))
US_regions <- read_csv(here("datasets", "US_regions.csv"))

popgrowth <- left_join(US_census, US_regions) %>%
    group_by(region, division, state) %>%
    summarize(
      pop2000 = sum(pop2000, na.rm = TRUE),
      pop2010 = sum(pop2010, na.rm = TRUE),
      popgrowth = (pop2010-pop2000)/pop2000
    ) %>%
    arrange(popgrowth) %>%
    ungroup() %>%
    mutate(
      #state = factor(state, levels = state),
      region = factor(region, levels = c("West", "South", "Midwest", "Northeast"))
    )

region_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442")

labeled_states <- c(
  "Alaska", "Arizona", "California", "Florida", "Wisconsin", 
  "Louisiana", "Nevada", "Michigan", "Montana", "New Mexico",
  "Pennsylvania", "New York", "Oregon", "Rhode Island",
  "Tennessee", "Texas", "Utah", "Vermont"
)

df_repel <- select(popgrowth, x = pop2000, y = popgrowth, state) %>%
  mutate(label = ifelse(state %in% labeled_states, as.character(state), ""))

p <- ggplot(popgrowth, aes(x = pop2000, y = popgrowth, color = region, fill = region)) +
  geom_text_repel(
    data = df_repel,
    aes(x, y, label = label),
    segment.alpha = 0.5, point.padding = 0.25,
    box.padding = .8,
    force = 1,
    min.segment.length = 0.1,
    max.overlaps = 1000,
    size = 10/.pt,
    seed = 7586,
    inherit.aes = FALSE
  ) +
  geom_point(size = 4, color = "white") +
  geom_point(size = 3, shape = 21) +
  scale_x_log10(
    breaks = c(1e6, 3e6, 1e7, 3e7),
    labels = expression(10^6, 3 %*% 10^6, 10^7, 3 %*% 10^7)
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab("population size in 2000") +
  ylab("population growth, 2000 to 2010") +
  theme_minimal_grid(14)

p +  
  scale_fill_manual(values = region_colors) +
  scale_color_manual(values = darken(region_colors, .3))

```
]

Palette name: Okabe-Ito

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Qualitative scale example

.center.nogap[
```{r popgrowth-vs-popsize-colored2, echo = FALSE, message = FALSE, out.width = "65%", fig.width = 7.5, fig.asp = 0.618, dev = "svg"}

region_colors <- RColorBrewer::brewer.pal(4, "Set1")

p +  
  scale_fill_manual(values = region_colors) +
  scale_color_manual(values = darken(region_colors, .3))

```
]

Palette name: ColorBrewer Set1

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Qualitative scale example

.center.nogap[
```{r popgrowth-vs-popsize-colored3, echo = FALSE, message = FALSE, out.width = "65%", fig.width = 7.5, fig.asp = 0.618, dev = "svg"}

region_colors <- RColorBrewer::brewer.pal(4, "Set3")

p +  
  scale_fill_manual(values = region_colors) +
  scale_color_manual(values = darken(region_colors, .3))

```
]

Palette name: ColorBrewer Set3

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>


---

# Sequential scale example

<br>

.center[
```{r four-locations-temps-by-month, echo = FALSE, message = FALSE, out.width = "90%", fig.width = 9, fig.asp = .3, dev = "svg"}
temps_months <- read_csv(here("datasets", "tempnormals.csv")) %>%
  group_by(location, month_name) %>%
  summarize(mean = mean(temperature)) %>%
  mutate(
    month = factor(
      month_name,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    ),
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(-month_name)

p <- ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = .95, height = 0.95) + 
  scale_y_discrete(name = NULL) +
  coord_fixed(expand = FALSE) +
  theme_half_open() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_text(size = 12)
  )

p + scale_fill_viridis_c(
    option = "D",
    name = "temperature (°F)"
  )
```
]

Palette name: Viridis

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Sequential scale example

<br>

.center[
```{r four-locations-temps-by-month2, echo = FALSE, message = FALSE, out.width = "90%", fig.width = 9, fig.asp = .3, dev = "svg"}
p + scale_fill_viridis_c(
    option = "B", begin = 0.15, end = 0.98,
    name = "temperature (°F)"
  )
```
]

Palette name: Inferno

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

# Sequential scale example

<br>

.center[
```{r four-locations-temps-by-month3, echo = FALSE, message = FALSE, out.width = "90%", fig.width = 9, fig.asp = .3, dev = "svg"}
p + scale_fill_viridis_c(
    option = "E",
    name = "temperature (°F)"
  )
```
]

Palette name: Cividis

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
3. Represent numeric values (diverging)
</td>
<td>
<img src = "color-scales_files/diverging.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

---

# Diverging scale example

.center.nogap[
```{r forensic-correlations1, echo = FALSE, message = FALSE, out.width = "35%", fig.width = 4., fig.asp = 1, dev = "svg"}
forensic_glass <- read_csv(here("datasets", "forensic_glass.csv"))

cm <- cor(select(forensic_glass, -type, -RI, -Si))
df_wide <- as.data.frame(cm)
df_long <- stack(df_wide)
names(df_long) <- c("cor", "var1")
df_long <- cbind(df_long, var2 = rep(rownames(cm), length(rownames(cm))))
clust <- hclust(as.dist(1-cm), method="average") 
levels <- clust$labels[clust$order]
df_long$var1 <- factor(df_long$var1, levels = levels)
df_long$var2 <- factor(df_long$var2, levels = levels)

p <- ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor)) + 
  geom_tile(color = "white", linewidth = 1) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0)) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"),
      ticks.linewidth = 1
    )
  ) +
  coord_fixed() +
  theme_half_open(rel_small = 1) +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = grid::unit(3, "pt"),
    legend.position = "inside",
    legend.position.inside = c(.97, .0),
    legend.justification = c(1, 0),
    legend.title = element_text(hjust = 0.5)
  )
p + scale_fill_distiller(
  name = "correlation",
  limits = c(-.5, .5),
  breaks = c(-.5, 0, .5),
  labels = c("–0.5", "0.0", "0.5"),
  type = "div",
  palette = "PiYG",
  direction = 1
)
```
]

Palette name: ColorBrewer PiYG

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Diverging scale example

.center.nogap[
```{r forensic-correlations2, echo = FALSE, message = FALSE, out.width = "35%", fig.width = 4., fig.asp = 1, dev = "svg"}
p + scale_fill_continuous_divergingx(
  name = "correlation",
  limits = c(-.5, .5),
  breaks = c(-.5, 0, .5),
  labels = c("–0.5", "0.0", "0.5"),
  palette = "Earth",
  rev = FALSE
)
```
]

Palette name: Carto Earth

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Diverging scale example

.center.nogap[
```{r forensic-correlations3, echo = FALSE, message = FALSE, out.width = "35%", fig.width = 4., fig.asp = 1, dev = "svg"}
p + scale_fill_continuous_diverging(
  name = "correlation",
  limits = c(-.5, .5),
  breaks = c(-.5, 0, .5),
  labels = c("–0.5", "0.0", "0.5"),
  palette = "Blue-Red",
  rev = TRUE
)
```
]

Palette name: Blue-Red

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle;"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
3. Represent numeric values (diverging)
</td>
<td>
<img src = "color-scales_files/diverging.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
4. Highlight
</td>
<td>
<img src = "color-scales_files/highlight.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

---

# Highlight example

.center[
```{r Aus-athletes-track, echo = FALSE, out.width = "60%", fig.width = 6, fig.asp = 0.618, dev = "svg"}
male_Aus <- ggridges::Aus_athletes %>%
  filter(sex == "m") %>%
  filter(
    sport %in% c("basketball", "field", "swimming", "track (400m)", "track (sprint)", "water polo")
  ) %>%
  mutate(
    sport = 
      case_when(
        sport == "track (400m)" ~ "track",
        sport == "track (sprint)" ~ "track",
        TRUE ~ sport
      ),
    sport = factor(
      sport,
      levels = c("track", "field", "water polo", "basketball", "swimming")
    )
  )

p <- ggplot(male_Aus, aes(x=height, y=pcBfat, shape=sport, color = sport, fill = sport)) +
  geom_point(size = 3) +
  scale_shape_manual(values = 21:25) +
  xlab("height (cm)") +
  ylab("% body fat") +
  theme_minimal_grid(14)

colors <- c("#BD3828", rep("#808080", 4))
fills <- c(
  alpha(colors[1], .815),
  alpha(colors[2:5], .5)
)

p + 
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills)
```
]

Palette name: Grays with accents

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Highlight example

.center[
```{r Aus-athletes-track2, echo = FALSE, out.width = "60%", fig.width = 6, fig.asp = 0.618, dev = "svg"}

accent_OkabeIto <- c("#E69F00", "#56B4E9", "#CC79A7", "#F0E442", "#0072B2", "#009E73", "#D55E00")
accent_OkabeIto[1:4] <- desaturate(lighten(accent_OkabeIto[1:4], .4), .8)
accent_OkabeIto[5:7] <- darken(accent_OkabeIto[5:7], .3)

colors <- c(accent_OkabeIto[5], darken(accent_OkabeIto[1:4], .2))
fills <- c(
  alpha(accent_OkabeIto[5], .7),
  alpha(accent_OkabeIto[1:4], .7)
)

p + 
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills)
```
]

Palette name: Okabe-Ito accent

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Highlight example

.center[
```{r Aus-athletes-track3, echo = FALSE, out.width = "60%", fig.width = 6, fig.asp = 0.618, dev = "svg"}

accent_Brewer <- RColorBrewer::brewer.pal(7, name = "Accent")[c(7, 1:4)]

colors <- darken(accent_Brewer, .2)
fills <- c(accent_Brewer[1], alpha(accent_Brewer[2:5], .7))

p + 
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills)
```
]

Palette name: ColorBrewer accent

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle;"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
3. Represent numeric values (diverging)
</td>
<td>
<img src = "color-scales_files/diverging.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
4. Highlight
</td>
<td>
<img src = "color-scales_files/highlight.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

---

class: center middle

## How does a computer represent color?

---

## The RGB color space

.pull-left.width-50[
- red R (0-255): amount of red light

- green G (0-255): amount of green light

- blue B (0-255): amount of blue light
]

--

.pull-right.move-down-1em[

R | G | B | hex code | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;color&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
---| --- | --- | --- | ---
0  | 0 | 0 | #000000 | <div style = "background-color: #000000;">&nbsp;</div>
255  | 0 | 0 | #FF0000 | <div style = "background-color: #FF0000;">&nbsp;</div>
0  | 255 | 255 | #00FFFF | <div style = "background-color: #00FFFF;">&nbsp;</div>
128  | 128 | 128 | #808080 | <div style = "background-color: #808080;">&nbsp;</div>
0 | 158 | 115 | #009E73 | <div style = "background-color: #009E73;">&nbsp;</div>
255 | 255 | 255 | #FFFFFF | <div style = "background-color: #FFFFFF;">&nbsp;</div>
]

---

## The RGB color space

.center[
```{r rgb-viz, echo = FALSE, cache = params$cache, out.width = "85%", fig.width = 10, fig.asp = .5, dev = "svg"}
rgb_by_b <- expand_grid(
  R = seq(0, 255, by = 2),
  G = seq(0, 255, by = 2),
  B = seq(0, 255, by = 35)
)

rgb_by_b$hex <- hex(sRGB(as.matrix(rgb_by_b)/255))

rgb_by_b %>%
  ggplot(aes(R, G, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "red R",
    breaks = (0:5)*50,
    labels = c("0", "", "100", "", "200", "")
  ) +
  scale_y_continuous(
    name = "green G",
    breaks = (0:5)*50,
    labels = c("0", "", "100", "", "200", "")
  ) +
  facet_wrap(
    vars(B),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("blue B = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace() +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]

--

.tiny[
People cannot reason well about the RGB color space
]

---

## The HSV color space

.pull-left[

- hue H (0-360):  
  hue of the color

- saturation S (0-1):  
  colorfulness relative to the brightness of the color

- value V (0-1):  
  subjective perception of amount of light emitted
]

.pull-right.move-up-1em[
```{r hsv-viz-single, echo = FALSE, cache = params$cache, out.width = "100%", fig.width = 5, fig.asp = 1, dev = "svg"}
hsv_by_h <- expand_grid(
  H = seq(0, 330, by = 30),
  S = seq(0, 1, by = .01),
  V = seq(0, 1, by = .01)
)

hsv_by_h$hex <- hex(HSV(as.matrix(hsv_by_h)))

hsv_by_h %>%
  filter(H == 90) %>%
  ggplot(aes(S, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "value V",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace(16) +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]


---

## The HSV color space

.center[
```{r hsv-viz, echo = FALSE, cache = params$cache, out.width = "85%", fig.width = 10, fig.asp = .5, dev = "svg"}
hsv_by_h %>%
  ggplot(aes(S, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "value V",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace() +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]

---

## The HLS color space

.pull-left[
- hue H (0-360):  
  hue of the color

- lightness L (0-1):  
  brightness relative to the brightness of a similarly illuminated white

- saturation S (0-1):  
  colorfulness relative to the brightness of the color
]

.pull-right.move-up-1em[
```{r hls-viz-single, echo = FALSE, cache = params$cache, out.width = "100%", fig.width = 5, fig.asp = 1, dev = "svg"}
hls_by_h <- expand_grid(
  H = seq(0, 330, by = 30),
  L = seq(0, 1, by = .01),
  S = seq(0, 1, by = .01)
)

hls_by_h$hex <- hex(HLS(as.matrix(hls_by_h)))

hls_by_h %>%
  filter(H == 90) %>%
  ggplot(aes(S, L, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "lightness L",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace(16) +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```

]

---

## The HLS color space

.center[
```{r hls-viz, echo = FALSE, cache = params$cache, out.width = "85%", fig.width = 10, fig.asp = .5, dev = "svg"}
hls_by_h %>%
  ggplot(aes(S, L, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "lightness L",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace() +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]


--

.tiny[
Reasoning about HSV or HLS can still be difficult
]

---

## The Luv color space

.pull-left.width-50[

- luminance L (0-100):  
  amount of light emitted

- green/red axis u (approx. -100 to 100):  
  amount of green vs. red

- blue/yellow axis v (approx. -100 to 100):  
  amount of blue vs. yellow
]

.pull-right.move-up-2em[
```{r luv-viz-single, echo = FALSE, cache = params$cache, out.width = "100%", fig.width = 5, fig.asp = 1, dev = "svg"}
luv_by_l <- expand_grid(
  L = seq(20, 90, by = 10),
  U = seq(-80, 170, by = 1),
  V = seq(-145, 110, by = 1),
)

luv_by_l$hex <- hex(LUV(as.matrix(luv_by_l)))

luv_by_l %>%
  filter(L == 60) %>%
  ggplot(aes(U, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_x_continuous(
    name = "green/red axis u",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_y_continuous(
    name = "blue/yellow axis v",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_fill_identity() +
  facet_wrap(
    vars(L),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("luminance L = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace(16)
```
]


---

## The Luv color space

.center[
```{r luv-viz, echo = FALSE, cache = params$cache, out.width = "85%", fig.width = 10, fig.asp = .5, dev = "svg"}
luv_by_l %>%
  ggplot(aes(U, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_x_continuous(
    name = "green/red axis u",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_y_continuous(
    name = "blue/yellow axis v",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_fill_identity() +
  facet_wrap(
    vars(L),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("luminance L = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace()
```
]

---

## The HCL color space: polar Luv

.center.move-up-1em[
```{r hcl-viz-single, echo = FALSE, cache = params$cache, out.width = "50%", fig.width = 5.5, fig.asp = 1, dev = "svg"}

rC <- 60
hue_df <- tibble(
  t = seq(0, 2*pi, length.out = 100),
  U = rC*sin(t),
  V = rC*cos(t)
)

chroma_df <- tibble(
  U = 4*rC*sin(pi/3),
  V = 4*rC*cos(pi/3)
)

labels_df <- tibble(
  U = c(1.4*rC*sin(pi/3), 1.01*rC*sin(.8*pi)),
  V = c(1.4*rC*cos(pi/3), 1.01*rC*cos(.8*pi)),
  label = c("chroma C", "hue L")
)

luv_by_l %>%
  filter(L == 60) %>%
  ggplot(aes(U, V)) +
  geom_raster(aes(fill = hex), na.rm = TRUE) +
  geom_segment(
    data = chroma_df,
    aes(x = 0, y = 0, xend = U, yend = V),
    linewidth = 0.2
  ) +
  geom_path(
    data = hue_df,
    linewidth = 0.2
  ) +
  geom_text(
    data = labels_df,
    aes(label = label),
    size = 16*(12/14)/.pt,
    hjust = 0, vjust = 1
  ) +
  scale_x_continuous(
    name = "green/red axis u",
    breaks = (-3:3)*50,
    labels = c("", "-100", "-50", "0", "50", "100", "")
  ) +
  scale_y_continuous(
    name = "blue/yellow axis v",
    breaks = (-3:3)*50,
    labels = c("", "-100", "-50", "0", "50", "100", "")
  ) +
  scale_fill_identity() +
  facet_wrap(
    vars(L),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("luminance L = {x}"))
  ) +
  coord_fixed(
    expand = FALSE,
    xlim = c(-65, 145),
    ylim = c(-120, 80)
  ) +
  theme_colorspace(16)
```
]


---

## The HCL color space: polar Luv

.move-up-1em.center[
```{r cl-planes, echo = FALSE, cache = params$cache, out.width = "80%", fig.width = 9, fig.asp = .6, dev = "svg"}
hcl_by_h <- expand_grid(
  L = seq(1, 99, by = 0.5),
  C = seq(0, 155, by = 1),
  H = seq(0, 330, by = 30)
)

hcl_by_h$hex <- hex(polarLUV(as.matrix(hcl_by_h)))

hcl_by_h %>%
  ggplot(aes(C, L, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "chroma C",
    breaks = (0:3)*50,
    labels = c("0", "", "100", "")
  ) +
  scale_y_continuous(
    name = "luminance L",
    limits = c(0, 100),
    breaks = (0:4)*25,
    labels = c("0", "", "50", "", "100")
  ) +
  facet_wrap(
    vars(H),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace()
```
]

[//]: # "segment ends here"


---
class: center middle

## Explore HCL colors interactively in R

---

## We can explore HCL colors in R

```{r eval = FALSE}
colorspace::choose_color()
```

--

<img src = "color-spaces_files/choose_color1.png", width = 35%, style = "position:absolute; top: 20%; left: 52.5%; box-shadow: 3px 5px 3px 1px #00000080;"></img>

--

<img src = "color-spaces_files/choose_color2.png", width = 35%, style = "position:absolute; top: 35%; left: 12.5%; box-shadow: 3px 5px 3px 1px #00000080;"></img>

--

<img src = "color-spaces_files/choose_color3.png", width = 35%, style = "position:absolute; top: 50%; left: 42.5%; box-shadow: 3px 5px 3px 1px #00000080;"></img>


[//]: # "segment ends here"

---
class: center middle

## A few considerations when choosing colors

---

## 1. Avoid high chroma

--

.pull-left.width-45[
High chroma: Toys

<img src = "color-spaces_files/toys-image.jpg", width = 100%, style = "box-shadow: 3px 5px 3px 1px #00000080;"></img>
]

???

[Photo by Pixabay from Pexels](https://www.pexels.com/photo/super-mario-and-yoshi-plastic-figure-163077/)

--

.pull-right.width-45[
Low chroma: Elegance

<img src = "color-spaces_files/home-interior.jpg", width = 95%, style = "box-shadow: 3px 5px 3px 1px #00000080;"></img>
]

???

[Photo by Saviesa Home from Pexels](https://www.pexels.com/photo/kitchen-island-2089698/)

---

## 2. Be aware of color-vision deficiency

5%–8% of men are color blind!

--

<img src = "dataviz/pitfalls_of_color_use_files/figure-html/red-green-cvd-sim-1.png", width = 100%></img>

Red-green color-vision deficiency is the most common

---

## 2. Be aware of color-vision deficiency

5%–8% of men are color blind!

<img src = "dataviz/pitfalls_of_color_use_files/figure-html/blue-green-cvd-sim-1.png", width = 100%></img>

Blue-green color-vision deficiency is rare but does occur

---

## 2. Be aware of color-vision deficiency

Choose colors that can be distinguished with CVD

<img src = "dataviz/pitfalls_of_color_use_files/figure-html/PiYG-cvd-sim-1.png", width = 100%></img>

---

## Consider using the Okabe-Ito scale for categorical data

.center[
<img src = "dataviz/pitfalls_of_color_use_files/figure-html/palette-Okabe-Ito-1.png", width = 100%></img>
]

.tiny-font[
Name           | Hex code &nbsp;&nbsp; | R, G, B (0-255)
:----------    | :-------  | :--------
orange         | #E69F00   | 230, 159, 0
sky blue	     | #56B4E9   | 86, 180, 233
bluish green   | #009E73   | 0, 158, 115
yellow	       | #F0E442   | 240, 228, 66
blue	         | #0072B2   | 0, 114, 178
vermilion	     | #D55E00   | 213, 94, 0
reddish purple | #CC79A7   | 204, 121, 167
black	         | #000000   | 0, 0, 0
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## CVD is worse for thin lines and tiny dots

.move-up-1em.center[
<img src = "dataviz/pitfalls_of_color_use_files/figure-html/colors-thin-lines-1.png", width = 65%></img>
]

---

## When in doubt, run CVD simulations

<div class = "move-up-1em" style = "width: 50%; box-shadow: 3px 5px 3px 1px #00000080; padding: 8px;">
.small-font[Original]<br>
<img src = "dataviz/redundant_coding_files/figure-html/iris-scatter-three-shapes-1.png", width = 100%></img>
</div>

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## When in doubt, run CVD simulations

<div class = "move-up-1em" style = "width: 50%; box-shadow: 3px 5px 3px 1px #00000080; padding: 8px;">
.small-font[Original]<br>
<img src = "dataviz/redundant_coding_files/figure-html/iris-scatter-three-shapes-1.png", width = 100%></img>
</div>

<img src = "dataviz/redundant_coding_files/figure-html/iris-scatter-three-shapes-cvd-1.png", width = 60% style = "position: absolute; top: 22%; left: 30%; box-shadow: 3px 5px 3px 1px #00000080; padding: 16px 8px 8px 16px; background: white"></img>

???

Figures from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"


---

## Further reading

- Fundamentals of Data Visualization: [Chapter 19: Common pitfalls of color use](https://clauswilke.com/dataviz/color-pitfalls.html)
- Wikipedia: [HSL and HSV](https://en.wikipedia.org/wiki/HSL_and_HSV)
- **colorspace** package documentation: [Color Spaces](https://colorspace.r-forge.r-project.org/articles/color_spaces.html)
- **colorspace** package documentation: [Apps for Choosing Colors and Palettes Interactively](https://colorspace.r-forge.r-project.org/articles/hclwizard.html)

