---
title: "Effective Data Visualization with ggplot2"
subtitle: "Sprucing up your legends"
format: html
editor: visual
---

## Required packages

Install the required packages:

```{r}
#| eval = FALSE

# Run this command to install the required packages.
# You need to do this only once.
install.packages(
  c(
    "tidyverse", 
  )
)
```

## 1. Legend order

By default, ggplot places legend entries in alphabetical order, and that is rarely what we want. See this example:

```{r}
#| message = FALSE,
#| warning = FALSE

library(tidyverse)

tech_stocks <- read_csv("https://wilkelab.org/dataviz_shortcourse/datasets/tech_stocks.csv") |>
  mutate(date = ymd(date)) |>
  select(company, date, price_indexed)

ggplot(tech_stocks) +
  aes(x = date, y = price_indexed) +
  geom_line(aes(color = company), na.rm = TRUE) +
  scale_x_date(
    limits = c(
      ymd("2012-06-01"),
      ymd("2017-05-31")
    ),
    expand = c(0, 0)
  ) + 
  scale_y_continuous(
    limits = c(0, 560),
    expand = c(0, 0)
  )
```

The visual order of the lines is Facebook, Alphabet, Microsoft, Apple, and we should match it by reodering the categorical variable.

```{r}
tech_stocks |>
  mutate(
    company = fct_relevel(
      company,
      "Facebook",  "Alphabet", "Microsoft", "Apple"
    )
  ) |>
  ggplot() +
  aes(x = date, y = price_indexed) +
  geom_line(aes(color = company), na.rm = TRUE) +
  scale_x_date(
    limits = c(
      ymd("2012-06-01"),
      ymd("2017-05-31")
    ),
    expand = c(0, 0)
  ) + 
  scale_y_continuous(
    limits = c(0, 560),
    expand = c(0, 0)
  )
```

This is improved, but we can do better. We can get rid of the legend entirely, by adding a secondary axis.

```{r}
tech_stocks_last <- tech_stocks |>
  filter(date == max(date))

ggplot(tech_stocks) +
  aes(x = date, y = price_indexed) +
  geom_line(aes(color = company), na.rm = TRUE) +
  scale_x_date(
    limits = c(
      ymd("2012-06-01"),
      ymd("2017-05-31")
    ),
    expand = c(0, 0)
  ) + 
  scale_y_continuous(
    limits = c(0, 560),
    expand = c(0, 0),
    sec.axis = dup_axis(
      breaks = tech_stocks_last$price_indexed,
      labels = tech_stocks_last$company,
      name = NULL
    )
  ) +
  guides(color = "none")
```

## 2. Legend placement
