---
title: "Effective Data Visualization with ggplot2"
subtitle: "Gradient and pattern fills"
format: html
editor: visual
knitr:
    opts_chunk: 
      dev: "ragg_png"
---

## Load required packages

```{r}
#| message = FALSE

library(tidyverse)
library(grid)
library(cowplot)
library(palmerpenguins)
library(sf)
```

## Solutions, Section 1

1.  Make your own gradient for a plot with gradient fill. Try both a linear gradient (made with `grid::linearGradient()`) and a radial gradient (made with `grid::radialGradient()`). Try different colors, gradient directions, and other modifications these functions allow you to make.

2.  Make your own patterns for the bar plot with pattern fill.

## Solutions, Section 2

Data preparation:

```{r}
#| message = FALSE

# data taken from: https://github.com/john-guerra/US_Elections_Results/tree/master
votes_2016 <- read_csv("https://wilkelab.org/dataviz_shortcourse/datasets/2016_US_County_Level_Presidential_Results.csv") |>
  mutate(
    fips = str_pad(combined_fips, 5, pad = "0"),
    state = state_abbr
  )
  
votes_long <- votes_2016 |>
  mutate(
    other = total_votes - votes_dem - votes_gop
  ) |>
  select(
    state, fips, democratic = votes_dem, republican = votes_gop, other
  ) |>
  pivot_longer(c(-state, -fips), names_to = "party", values_to = "votes")

# load geometry data
counties <- readRDS(url("https://wilkelab.org/dataviz_shortcourse/datasets/US_counties.rds")) |>
  mutate(
    fips = as.character(GEOID),
    state = state_code
  )
```

Use side-by-side bar plots instead of pie charts.

Solution: First we generate the patterns.

```{r}
make_bar_plot <- function(data) {
  data |>
    mutate(
      party = fct_relevel(party, 'republican', 'democratic', 'other')
    ) |>
    ggplot() +
    aes("", votes, fill = party) +
    geom_col(color = "black", linewidth = 0.2, position = "dodge") +
    scale_fill_manual(
      values = c(democratic = '#4B77D0', republican = '#DB4940', other = "gray70"),
      guide = "none"
    ) +
    theme_void()
}

make_bar_pattern <- function(data) {
  p <- make_bar_plot(data)
  pattern(
    ggplotGrob(p),
    extend = "none",
    group = FALSE
  )
}

vote_bars <- votes_long |>
  filter(state == "WY") |>
  select(-state) |>
  nest(data = -fips) |>
  mutate(
    bar_pattern = map(data, make_bar_pattern),
    vote_total = map_dbl(data, ~sum(.x$votes))
  ) |>
  select(-data)
```

Then we make the plot.

```{r}
counties |>
  filter(state == "WY") |>
  mutate( # calculate reference point for each county
    points = st_point_on_surface(st_zm(geometry)),
    county_x = st_coordinates(points)[, "X"],
    county_y = st_coordinates(points)[, "Y"]
  ) |>
  left_join(vote_bars, by = "fips") |>
  mutate( # calculate plot scale
    scale = 1.2 * vote_total,
  ) |>
  ggplot() +
  geom_sf(color = "gray40", fill = "gray95", linewidth = 0.2) +
  geom_rect(
    aes(
      geometry = geometry,
      xmin = county_x - 25000,
      xmax = county_x + 25000,
      ymin = county_y - .5 * scale,
      ymax = county_y + 1.5 * scale,
      fill = bar_pattern
    ),
    stat = "sf_coordinates"
  ) +
  theme_void()
```

## Solutions, Section 3

1.  Create a version of the pet ownership isotype plot where bars run vertical instead of horizontal.

```{r}
make_img_pattern <- function(img) {
  pattern(
    rasterGrob(
      magick::image_read(img),
      width = unit(10, "mm"),
      height = unit(10, "mm"),
      interpolate = FALSE,
      vp = viewport(
        x = 0,
        just = "left"
      )
    ),
    width = unit(100, "mm"),
    height = unit(10, "mm"),
    extend = "repeat",
    group = FALSE
  )
}

patterns <- map(
  c(
    "https://wilkelab.org/dataviz_shortcourse/images/dog.png",
    "https://wilkelab.org/dataviz_shortcourse/images/cat.png",
    "https://wilkelab.org/dataviz_shortcourse/images/fish.png",
    "https://wilkelab.org/dataviz_shortcourse/images/bird.png"
  ),
  make_img_pattern
)

# data source: 2012 U.S. Pet Ownership & Demographics Sourcebook,
# American Veterinary Medical Association
pet_ownership <- read.table(text = "pet  households
dogs 43346000
cats 36117000
fish	7738000
birds 3671000
", header = TRUE)

pet_ownership |>
  mutate(
    pet = fct_reorder(pet, -households)
  ) |>
  ggplot() +
  aes(x = pet, y = households, fill = pet) +
  geom_col() +
  geom_label(
    aes(label = paste0(signif(households*1e-6, 2), "M")),
    hjust = 0.5,
    vjust = 0,
    nudge_y = .1e6,
    size = 14,
    size.unit = "pt",
    label.size = 0, # no label outline
    label.padding = unit(2, "pt"),
    fill = "#FFFFFF"
  ) +
  scale_y_continuous(
    limits = 1e6*c(0, 49),
    breaks = 1e7*(0:4),
    labels = c("0", paste0(10*(1:4), "M")),
    name = "households",
    expand = c(0, 0)
  ) +
  scale_x_discrete(
    name = NULL
  ) +
  scale_fill_manual(
    values = patterns,
    guide = "none"
  ) +
  theme_minimal_hgrid(rel_small = 1)
```

2.  Recreate the pet ownership isotype plot with ggpattern.

```{r}
#| warning = FALSE
library(ggpattern)

pet_images <- c(
  "https://wilkelab.org/dataviz_shortcourse/images/bird.png",
  "https://wilkelab.org/dataviz_shortcourse/images/fish.png",
  "https://wilkelab.org/dataviz_shortcourse/images/cat.png",
  "https://wilkelab.org/dataviz_shortcourse/images/dog.png"
)

pet_ownership |>
  mutate(
    pet = fct_reorder(pet, households)
  ) |>
  ggplot() +
  aes(y = pet, x = households) +
  geom_col_pattern(
    aes(
      pattern_filename = pet
    ), 
    pattern         = 'image',
    pattern_type    = 'tile',
    fill            = 'white', 
    colour          = NA,
    pattern_filter  = 'box',
    pattern_scale   = -2
  ) +
  geom_label(
    aes(label = paste0(signif(households*1e-6, 2), "M")),
    vjust = 0.5,
    hjust = 0,
    nudge_x = .1e6,
    size = 14,
    size.unit = "pt",
    label.size = 0, # no label outline
    label.padding = unit(2, "pt"),
    fill = "#FFFFFF"
  ) +
  scale_x_continuous(
    limits = 1e6*c(0, 49),
    breaks = 1e7*(0:4),
    labels = c("0", paste0(10*(1:4), "M")),
    name = "households",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL
  ) +
  scale_pattern_filename_discrete(
    choices = pet_images,
    guide = "none"
  ) +
  theme_minimal_hgrid(rel_small = 1)
```
